%!TEX root = ../thesis_a4.tex
\chapter{Appendix A: Publications}
\label{sec:pubs}

In this appendix entry we list the peer-reviewed publications that are part of this thesis. Under each paper citation, a brief description of the contributions made by the author of this thesis is also provided. The articles are presented in chronological order.



\section*{Journal articles}

\begin{mybox}{JOSS, Journal of Open Software}
    Magdalena Fuentes, Gen\'{i}s Plaja-Roglans, Guillem Cort\`{e}s-Sebasti\`{a}, Tanmay Khandelwal, Marius Miron, Xavier Serra, Juan Pablo Bello, \& Justin Salamon. (2024). \emph{Soundata: Reproducible use of audio datasets.} Journal of Open Source Software. %9(98), 6634.
\end{mybox}
\textbf{Contributions:} Writing and proof-reading the manuscript, and working on improving the code and documentation, following the concerns and suggestions raised by the reviewers.
%
Also, the author of this thesis has been actively maintaining and upgrading the library from Fall 2021.


\section*{Conference papers}

\subsection*{2021}

\begin{mybox}{CMMR 2021}
    Thomas Nutall, Genís Plaja-Roglans, Lara Pearson \& Xavier Serra (2021).
    \emph{The Matrix profile for motif discovery in audio-an example application in Carnatic music}.
    In: Proc. of the 15th Int. Symp. on Computer Music Multidisciplinary Research (CMMR), Tokyo, Japan. %pp. 109-118.
\end{mybox}
\textbf{Contributions:} Writing and proof-reading the manuscript. Moreover, working on the pitch extraction and singing voice cleaning part of the pipeline proposed in the paper.


\subsection*{2022}

\begin{mybox}{FRSM 2022}
    Genís Plaja-Roglans, Thomas Nuttall, Xavier Serra \& Marius Miron (2022).
    \emph{Continuing CompMusic: New Approaches in the computational analysis of Carnatic music}.
    In: Proc. of 27th Int. Symposium on Frontiers of Research in Speech and Music (FRSM), Online.\footnote{Also published in Advances in Intelligent Systems and Computing Journal}
\end{mybox}
\textbf{Contributions:} Writing and proof-reading, and performing the pitch extraction and vocal separation experiments. 


\begin{mybox}{ISMIR 2022 - Main}
    Genís Plaja-Roglans, Marius Miron \& Xavier Serra (2022).
    \emph{A diffusion-inspired training strategy for singing voice extraction in the waveform domain}.
    In: Proc. of the 23rd Int. Society for Music Information Retrieval Conf.  (ISMIR), Bengaluru, India.
\end{mybox}
\textbf{Contributions:} Leading the work, performing all the experiments in the paper, and leading the writing and proof-reading of the document.

\begin{mybox}{ISMIR 2022 - Collaboration}
    Thomas Nutall, Genís Plaja-Roglans, Lara Pearson \& Xavier Serra (2022).
    \emph{In search of Sañcāras: tradition-informed repeated melodic pattern recognition in carnatic music}.
    In: Proc. of the 23rd Int. Society for Music Information Retrieval Conf. (ISMIR), Bengaluru, India.  % (pp. 337-344).
\end{mybox}
\textbf{Contributions:} In charge of the pitch extraction and vocal cleaning parts of the approach. Writing and proof-reading.


\subsection*{2023}

\begin{mybox}{ISMIR 2023}
    Genís Plaja-Roglans, Adithi Shankar, Marius Miron \& Xavier Serra (2023).
    \emph{Carnatic singing voice separation using cold diffusion on training data with bleeding}.
    In: Proc. of the 24rd Int. Society for Music Information Retrieval Conf. (ISMIR), Milan, Italy.
\end{mybox}
\textbf{Contributions:} Leading the work, performing all the training and evaluation experiments in the paper, and leading the writing and proof-reading of the document.

\newpage

\begin{mybox}{TISMIR 2023}
    Genís Plaja-Roglans, Thomas Nuttall, Lara Pearson, Xavier Serra \& Marius Miron (2023).
    \emph{Repertoire-specific vocal pitch data generation for improved melodic analysis of Carnatic Music}.
    Transactions of the Int. Society for Music Information Retrieval (TISMIR), 6(1).  %, p. 13-26.
\end{mybox}
\textbf{Contributions:} Leading the work, performing the data synthesis and pitch extraction experiments, writing and proof-reading.


\subsection*{2024}

\begin{mybox}{ISMIR 2024}
    Adithi Shankar, Genís Plaja-Roglans, Thomas Nuttall, Martín Rocamora \& Xavier Serra (2024).
    \emph{Saraga Audiovisual: a large multimodal open data collection for the analysis of carnatic music.}
    In: Proc. of the 25rd Int. Society for Music Information Retrieval Conf. (ISMIR), San Francisco, USA.
\end{mybox}
\textbf{Contributions:} Active participation on discussions, evaluation experiments, writing and proof-reading the manuscript.


\subsection*{2025}

\begin{mybox}{WIMAGA 2025}
    Adithi Shankar, Serafin Schweinitz, Genís Plaja-Roglans, Xavier Serra \& Martín Rocamora (2025).
    \emph{Disentangling overlapping sources: improving vocal and violin source separation in Carnatic Music.}
    In: Workshop on Indian Music Analysis and Generative Applications (WIMAGA) in ICASSP 2025, Hyderabad, India.
\end{mybox}
\textbf{Contributions:} Active participation on discussions, coding assistance, writing and proof-reading the manuscript.

\begin{mybox}{WASPAA 2025}
    Genís Plaja-Roglans, Yun-ning Hung, Xavier Serra \& Igor Pereira (2025).
    \emph{Generating separated singing vocals using a diffusion model conditioned on music mixtures.}
    In: Proc. of the IEEE Workshop on Applications of Signal Processing to Audio and Acoustics (WASPAA), Lake Tahoe, USA.
\end{mybox}
\textbf{Contributions:} Leading the work, coding the system, performing the training and evaluation experiments, writing and proof-reading.

\newpage

\begin{mybox}{IJCNN 2025}
    Genís Plaja-Roglans, Yun-ning Hung, Xavier Serra \& Igor Pereira (2025).
    \emph{Efficient and fast generative-based singing voice separation using a Latent Diffusion Model.}
    In: Proc. of the Int. Joint Conference of Neural Networks (IJCNN), Rome, Italy.
\end{mybox}
\textbf{Contributions:} Leading the work, coding the system, performing the training and evaluation experiments, writing and proof-reading.

\begin{mybox}{ISMIR 2025}
    Genís Plaja-Roglans, Xavier Serra \& Martín Rocamora (2025).
    \emph{Leveraging Carnatic live recordings for singing voice separation using regression-guided Latent Diffusion.}
    In: Proc. of the 25th Int. Society for Music Information Retrieval Conf. (ISMIR), Daejeon, Korea.
\end{mybox}
\textbf{Contributions:} Leading the work, coding the system, performing the training and evaluation experiments, writing and proof-reading.


\subsection*{Under review}

\begin{mybox}{audiovisual-bleed}
    Adithi Shankar, Genís Plaja-Roglans, Gloria Haro, Xavier Serra \& Martín Rocamora (-).
    \emph{Gesture-guided melodic source separation in Carnatic Music using cross-modal fusion techniques}
\end{mybox}
\textbf{Contributions:} Active participation on discussions, coding assistance, writing and proof-reading the manuscript.



\chapter{Appendix B: Other scientific outputs of this thesis}
\textbf{From Chapter~\ref{ch02_sb}}
\begin{itemize}
    \item ISMIR Tutorial: \url{https://mtg.github.io/IAM-tutorial-ismir22}
\end{itemize}
%\textbf{Contributions:} Leading the work, contribution on the presentation, coding and maintaining the published webbook. During ISMIR 2022, the author of this thesis presented one of the three main sections of the tutorial.


\textbf{From Chapter~\ref{ch03_data}}
\begin{itemize}
    \item Canonical Saraga: \url{https://zenodo.org/records/4301737}, unified access to the Dunya corpora: \url{https://mtg.github.io/compIAM/source/datasets.html\#access-the-dunya-corpora} (at the moment only for the \acrshort{IAM} repertoires.)
    \item New dataloaders (a total of $10$ for the computational analysis of Carnatic and Hindustani music), core improvements, and bug fixes in \texttt{mirdata} and \texttt{soundata}. Contributions led by the candidate of this thesis can be listed accessing the corresponding repos and filtering pulls by author:
    \begin{itemize}
        \item {\small \url{mir-dataset-loaders/mirdata/pulls?q=author%3Agenisplaja}}
        \item {\small \url{soundata/soundata/pulls?q=author%3Agenisplaja}}
    \end{itemize}
\end{itemize}


\textbf{From Chapter~\ref{ch04_melody}}
\begin{itemize}
    \item SCMS dataset: \url{https://zenodo.org/records/5553925}
    \item Code: \url{https://github.com/MTG/carnatic-pitch-patterns}
    \item FTANet-Carnatic: code and weights available in \texttt{compIAM}
\end{itemize}


\textbf{From Chapter~\ref{ch05_sep_nongen}}
\begin{itemize}
    \item Code for the iterative accompaniment removal model on musdb18hq is available at: \url{https://github.com/genisplaja/diffusion-vocal-sep}
    \item Code for bleeding-aware vocal separation system for Saraga is available at: \url{https://github.com/MTG/carnatic-separation-ismir23}
\end{itemize}


\textbf{From Chapter~\ref{ch06_sep_gen}}
\begin{itemize}
    \item Code to build, train and run inference with \textit{diff-dmx} and \textit{ldm-dmx} is available in \url{https://github.com/weAreMusicAI/dmx-diffusion}
    \item Code to build, train and run inference with \textit{ldm-dmx-bg} is available in \url{https://github.com/genisplaja/ldm-carnatic-separation}
\end{itemize}


\textbf{From Chapter~\ref{ch07_misc}}
\begin{itemize}
    \item Code for the unified framework including access to datasets, tools, and models for the computational analysis of Carnatic and Hindustani music: \url{https://github.com/MTG/compIAM}
    \item Code base for the diffusion experiment framework developed at Moises.ai: \url{https://github.com/weAreMusicAI/dmx-diffusion}
\end{itemize}
