Voldria agrair abans de tot i de tothom a la mare. Sense la teva perseverança no m'hauria matriculat a enginyeria, i menys encara haver-ho fet fins a tres cops! Quan el pare ens va deixar, no em vaig veure amb cor de fer primer de carrera a la UPF i ho vaig deixar, però poc després em vas animar a tornar a començar els estudis per internet (per la UOC) i vaig arribar a tercer. Quan semblava encarrilat, la UOC va decidir que no érem prou estudiants matriculats a la branca d'audiovisuals i que hauríem de fer informàtica, i malgrat que la UPF només va accedir a convalidar la meitat dels crèdits, no em vas deixar defallir i vam començar de nou. A la tercera va la vençuda, que diuen, oi? Sense tu potser hauria hagut de seguir descarregant camions a altes hores de la matinada, plens d'altaveus i de per vida.
  
Looking back, my sincere gratitude to Franck Giron, the inventor of the Monopole Synthesis model and my first mentor at Sony Stuttgart. His willingness to teach and counsel was very important, especially at such an early stage of my career. I keep remembering his advice: \textit{"listen to the actual thing!"}. Thanks also to Marc Ferras for his friendship at that time; Stuttgart would have been much worse without you. I hope we see each other at the slopes again one day, Marc.
 
Thanks to Jordi Pons, my second mentor, for teaching me how to do research and write papers. Without him I would have never been able to get into the research world. I am not always sure if I will end up doing research in the future, but I am grateful for having had the opportunity to do it until now. I remember calling you during the pandemic, when my main concern was that I could not buy a car! Thanks also for counseling me during the years of this PhD, and for introducing me to Joan Serrà and Santiago Pascual too, which helped a lot during that first paper.

I would also like to thank Xavier Serra, for having built something as cool and as great as the Music Technology Group precisely in Catalonia, and for the patience and support during the years of this work. Thanks to his understanding when I had to combine this work with my other professional commitments. Xavier, a vegades em venen pensaments intrusius recordant de les vegades que vaig demanar reunir-me amb tu per res, quan tenia 20 anys i em preocupaven minúcies.  I hope I have been able to repay that patience with some actual work. Thanks to Frederic for his feedback during the months of the PhD proposal; Freesound is amazing. Thanks to Dmitry Bogdanov for teaching me lots of stuff and what code reviews are.

My gratitude also to Umut Sayin, first and foremost for giving me the position and the chance to do a PhD. Thanks for all the meetings and counseling, as well as for setting the scope of the topics for Chapters \ref{ch03_hearing} and \ref{ch04_mbrirs}. Thanks for letting me steer the direction of the research a bit towards Music Information Retrieval (MIR). Thanks also for always having a smile on your face, even when I was in panic mode riding the roller coaster under lots of pressure. 


I also want to acknowledge the help of Joanna Luberadzka, my main collaborator, for her invaluable work and more importantly for her work ethic. It has been a real privilege to work with you, Joanna, and I hope we can keep doing it in the future. Thanks also to Martí --- I hope your PhD goes well! It was really nice to collaborate on Chapter 3. 

Many thanks to the rest of the team at Eurecat: Tim, Julien, Rafa, Roger and Adan... and to the rest of the MTG: Genís, Guillem, Roser, Tom, Pedro, Pablos... and many more. I have learned a lot from all of you, and I am grateful for the good times we have shared. 

Lastly, I want to thank my friends at La Sotana and from the bands I have been working with, for understanding when I had to cancel plans with them because of this work. Andreu, Joel, Magí i Manel ara tindreu tot un Dr. baixant-vos els vídeos de YouTube. I que duri! Joan, Aleix: gràcies per permetre'm vibrar amb vosaltres fent El Rock, i per deixar-me mesurar el PaRIRset durant la vostra gira. I a tots els altres, gràcies per ser-hi i per anar fent.

I l'agraïment més important és per la Gemma; tu has impedit que llencés la tovallola a la primera de canvi. Quina sort conèixer-te i comptar amb tu. T'estimo!
