\vspace*{\fill}

\begin{center}
\rule{4cm}{0.2pt}
\end{center}

This thesis has been carried out at the Music Technology Group (MTG) from the
Universitat Pompeu Fabra in Barcelona, Catalunya, Spain, from September 2021 to August 2025.
It was supervised by Prof. Xavier Serra Casals, and co-supervised by Dr. Marius Miron (from September 2021 to August 2023) and Dr. Martín Rocamora (from September 2023 to August 2025).
The work from Chapter 3 was conducted in collaboration with Dr. Magdalena Fuentes from the New York University (NYU), Dr. Guillem Cortès-Sebastià from BMAT Licensing S.L. in Barcelona, Tanmay Khandelwal (NYU), and several collaborators from both the MTG and NYU. It also involved several researchers that had been part of the CompMusic project, and undergraduate students from the Sound and Music Computing master's degree at UPF.
The work from Chapter 4 was conducted in collaboration with Dr. Lara Pearson, who is a researcher at Universität zu Köln, and Thomas Nuttall, a PhD fellow at the MTG.
The work from Chapter 5 was conducted in collaboration with Adithi Shankar, a PhD fellow at the MTG.
The work from Chapter 6 was conducted partly in a research internship at Moises.ai under the supervision of Dr. Igor Gadelha Pereira, and in collaboration with Dr. Yun-ning Hung, also member of Moises.ai. The second part of the work in Chapter 6 was conducted at the MTG under the supervision of Dr. Martín Rocamora and Prof. Xavier Serra.

Within the context of this tesis, there have been research efforts to contribute to work led by Adithi Shankar and Thomas Nuttall, PhD fellows at the MTG.

This research is part of \emph{MusicalAI}, funded by the Ministry of Science and Innovation of the Spanish Government and the Agencia Estatal de Investigación (AEI), with reference: PID2019-111403GB-I00 (2020-2024). This work was supported by the associated grant with reference: PRE2020-093110.